---
title: "Time Series Home work 3"
author: "Avery Loftin"
date: "9/26/2018"
output: pdf_document
---

## Question 1
(a) No, the expected value must be independent of t

(b) $E[Y_t]=E[7-3t+X_t]=7-3t+3t=7$

$\newline Var(Y_t)=Var(7-3t+X_t)=Var(X_t)=Cov(X_t,X_t)=\gamma _0$

$Cov(Y_t,Y_{t-k})=Cov(7-3t+X_t,7-3t+X_{t-k})=\gamma_k$

Yes, $\gamma_t$ is stationary because mean, variance, and autocovariance are all independent of t

(c)
```{r}
set.seed(2018)
time <- 1:100
noise <- rnorm(100, mean=0, sd=10)
X_t <- c()

for (t in time) {
  X_t[t] <- 3*t + noise[t]
}

plot.ts(X_t)
Y_t <- c()
for (t in time){
  Y_t[t] <- 7 - 3*t + X_t[t]
}

plot.ts(Y_t)
```

Simulation is consistent with analytical solution.

## Question 2

(a) odd $t$ and $k$
$cov(X_t, X_{t-k} + 3) = cov(X_t, X_{t-k}) = \gamma _k$

odd $k$ even $t$ or odd $t$ even $k$
$Cov(X_t, X_{t-k}) = \gamma _k$
even $t$ and $k$
$Cov(X_t + 3, X_{t-k} + 3) = cov(X_t, X_{t-k}) = \gamma _k$

(b)
$t$ is odd
$E[Y_t] = E[X_t] = constant$
$t$ is odd
$E[Yt]$ = $E[Xt+3] = 3 + constant$

$Var(Y_t) = Var(X_t + 3) = Var(X_t) = constant$

Yes, $Y_t$ is stationary because mean, variance, and autocovariance are all independent of $t$
```{r}
X_t <- c()
for (t in time) {
  X_t[t] <- rnorm(1)
}

Y_t <- c()
for (t in time) {
  if (t %% 2 == 0) {
    Y_t[t] <- X_t[t]
  }
  else {
    Y_t[t] <- X_t[t] + 3
  }
}

plot.ts(X_t)
plot.ts(Y_t)
```

Simulation shows $Y_t$ is indeed stationary using $X_t$~$N(0,1)$

## Question 3

(a) $E[Y_t] = \int{Y_t}d\varphi = \int cos(2\pi(\frac{t}{12} + \varphi) d\varphi = \frac{1}{2}\pi sin[2\pi(\frac{t}{12})+\varphi]$ evaluated from $0$ to $1$ $= \frac{1}{2}\pi [sin(2\pi(\frac{t}{12}+1)) - sin(2\pi(\frac{t}{12}))] = \frac{1}{2}\pi[0] = 0$

$Var[Y_t] = E[{Y_t}^2] - (E[Y_t])^2 = E[Y_t^2] = \int cos^2(2\pi(\frac{t}{12}+\varphi))d\varphi= \frac{1}{2}\int cos(4\pi(\frac{t}{12}+\varphi))+1d\varphi = \frac{1}{2}[\frac{1}{4\pi} sin(4\pi(\frac{t}{12}+\varphi)) + 1]$ evaluated from $0$ to $1$$= \frac{1}{8}\pi[sin(4\pi(\frac{t}{12}+1)) - sin(4\pi(\frac{t}{12}))] + \frac{1}{2} = \frac{1}{2}$

(b)
```{r}
Y_t <- c()
noise <- runif(100, min = 0, max = 1)
for (t in time) {
  Y_t[t] <- cos(2*pi*(t/12 + noise[t]))
}

plot.ts(Y_t)
```

$Y_t$ does not show any trend, as it is stationary.

